Model Checking Differentially Private Properties

### 2018-01-30 applications.tex
fix typo
proof appendix
work through and 16 pages ;)

Streaming processing gives a motivating example for Markov
decision processes. In stream processing, input data are retrieved and
analyzed continuously. Such continual observation is commonly found
in social media, service providers, or even electronic surveillance
systems. To protect privacy, privacy-respecting mechanisms must be
formulated and designed. Classical differential privacy however is
insufficient in streaming processing. Note that observations are of
indefinite lengths in such applications. All differentially private
mechanism necessarily stay in their internal states indefinitely
before outcomes are produced. Although differential privacy demands
privacy-respecting outcomes, internal states may reveal private
information in such mechanisms. When differentially private mechanisms
are in their internal states, privacy is still lost if an intrusion
occurs.

In pan privacy, data analysis mechanisms are required to respect
privacy even in their internal states~\cite{DNPR:10:DPCO,DNPRY:10:PPSA}.
The idea is to show reachability probabilities are similar on
neighboring observations. In the following, we model a discrete
variant of the mechanisms by Markov decision processes. We will also
specify pan privacy in our model using $\dpCTLstar$.

@@

In order to obtain a finite model, we formalize neighbors as action
sequences in Markov decision processes. Consider the action set $\{
a[a], a[b], b[a], b[b] \}$. Informally, the action $x[y]$ means the
event type $x$ is replaced by the event type $y$ in a neighbor. An
action sequence thus represents an observation and one of its
neighbors and vice versa. More formally, the action sequence
$x_1[y_1]x_2[y_2]\cdots x_n[y_n]$ represents the observation
$x_1x_2\cdots x_n$ and its neighbor $y_1y_2\cdots y_n$. For instance,
the action sequence $a[a]a[b]a[a]b[a]b[b]$ represents the observation
$aaabb$ and a neighbor $abaab$.

Similar to Algorithm~\ref{algorithm:frequency-estimator}, our Markov
decision process has internal states $s_{00}$, $s_{01}$, $s_{10}$, 
$s_{11}$ for the two bits $f_a$ and $f_b$ of observed event types
(Figure~\ref{figure:adjacent-inputs}). The internal state $s_{01}$
represents $(f_a, f_b) = (0, 1)$, for instance. Moreover, let $xy \in
L (s_{xy})$ for $x, y \in \{ 0, 1 \}$. From the
initialization state $s$, the Markov decision process moves to one of
the four internal states randomly with the uniform
distribution. Consider the internal state $s_{01}$. If the actions $a[a]$
or $a[b]$ are chosen, they denote that the event type $a$ is currently
observed. The Markov decision process moves to the internal state $s_{01}$
and $s_{11}$ with probability $.375 = \frac{3}{8}$ and $.625 =
\frac{5}{8}$ respectively. Other transitions are defined similarly.

On the other hand, the initialization state $t$
corresponds to the computation on neighbors. The Markov decision
process moves to the interal states $t_{00}$,
$t_{01}$, $t_{10}$, $t_{11}$ randomly with
the uniform distribution. Similarly, $xy \in L (t_{xy})$ for $x, y \in
\{ 0, 1 \}$.
Consider, for example, the internal state
$t_{10}$. If the action $a[a]$ or $b[a]$ are chosen, they
denote that the current event type is replaced with the eveny type
$a$. The Markov decision process moves to the internal state
$t_{00}$ and $t_{10}$ with probability $.375 =
\frac{3}{8}$ and $.625 = \frac{5}{8}$ respectively. At an internal
state, the Markov decision process can move to the output states $0$,
$1$, or $2$ by the truncated $\frac{1}{2}$-geometric mechanism over
$\{ 0, 1, 2 \}$ (Figure~\ref{figure:randomized-frequency}, \`{a} la
Figure~\ref{figure:geometric-mechanism}). For any action sequence,
observe that paths from the state $s$ and those from $t$
formalize the computation of Algorithm~\ref{algorithm:frequency-estimator}
on an observation and its neighbor represented by the sequence. We
hence say the states $s$ and $t$ are neighbors in the
Markov decision process (Figure~\ref{figure:frequency-estimator}).

\begin{figure}
  \centering
  \begin{subfigure}{\columnwidth}
    \centering
      \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node
      distance=2cm,node/.style={circle,draw}]
      \node[node] (I) at  ( 0,  0) { $s$ };
      \node[node] (00) at (-1.5,  1.5) { $s_{00}$ };
      \node at (-1.9, 1.9) { $00$ };
      \node[node] (01) at (-1.5, -1.5) { $s_{01}$ };
      \node at (-1.9, -1.9) { $01$ };
      \node[node] (10) at ( 1.5,  1.5) { $s_{10}$ };
      \node at ( 1.9,  1.9) { $10$ };
      \node[node] (11) at ( 1.5, -1.5) { $s_{11}$ };
      \node at ( 1.9, -1.9) { $11$ };

@@

@@

      \end{tikzpicture}
      \caption{Randomized Frequency}
      \label{figure:randomized-frequency}
    \end{subfigure}
  \caption{Frequency Estimator}
  \label{figure:frequency-estimator}
\end{figure}

\noindent
\emph{Privacy Specification}.
For classical differential privacy, we are interested in the output
distribution. Algorithm~\ref{algorithm:frequency-estimator} has three
different outputs. By privacy analysis, the state $I$ satisfies the
following $\dpCTLstar$ formula:
\[
\Psi = \dpriv{\ln 4}{0} (\F \mathit{at}_0) \wedge \dpriv{\ln 4}{0} (\F
\mathit{at}_1) \wedge \dpriv{\ln 4}{0} (\F \mathit{at}_2)p
\]
For pan privacy, we moreover wish to measure privacy loss in internal
states. For observations of length $k$, the probabilities of reaching
each internal states need to be sufficiently close. The state $I$
therefore must satisfy
\[
\Psi
\wedge
\bigwedge_{k \in \bbfN}
\dpriv{\ln 4}{0} \X^k {00} \wedge
\dpriv{\ln 4}{0} \X^k {01} \wedge
\dpriv{\ln 4}{0} \X^k {10} \wedge
\dpriv{\ln 4}{0} \X^k {11}.
\]

>>>>

Streaming processing gives a motivating example for MDPs. In stream 
processing, input data are retrieved and
analyzed continuously. Such continual observation is commonly found
in social media, service providers, or even electronic surveillance
systems. To protect privacy, privacy-respecting mechanisms must be
formulated and designed. Classical differential privacy however is
insufficient in streaming processing. Note that observations are of
indefinite lengths in such applications. All differentially private
mechanism necessarily stay in their internal states indefinitely
before outcomes are produced. Although differential privacy demands
privacy-respecting outcomes, internal states may reveal private
information in such mechanisms. When differentially private mechanisms
are in their internal states, privacy is still lost if an intrusion
occurs.

In pan privacy, data analysis mechanisms are required to respect
privacy even in their internal states~\cite{DNPR:10:DPCO,DNPRY:10:PPSA}.
The idea is to show reachability probabilities are similar on
neighboring observations. In the following, we model a discrete
variant of the mechanisms by MDPs. We will also
specify pan privacy in our model using $\dpCTLstar$.

@@

In order to obtain a finite model, we formalize neighbors as action
sequences in MDPs. Consider the action set $\{
a[a], a[b], b[a], b[b] \}$. Informally, the action $x[y]$ means the
event type $x$ is replaced by the event type $y$ in a neighbor. An
action sequence thus represents an observation and one of its
neighbors and vice versa. More formally, the action sequence
$x_1[y_1]x_2[y_2]\cdots x_n[y_n]$ represents the observation
$x_1x_2\cdots x_n$ and its neighbor $y_1y_2\cdots y_n$. For instance,
the action sequence $a[a]a[b]a[a]b[a]b[b]$ represents the observation
$aaabb$ and a neighbor $abaab$.

Similar to Algorithm~\ref{algorithm:frequency-estimator}, our MDP has internal states $s_{00}$, $s_{01}$, $s_{10}$,
$s_{11}$ for the two bits $f_a$ and $f_b$ of observed event types
(Figure~\ref{figure:adjacent-inputs}). The internal state $s_{01}$
represents $(f_a, f_b) = (0, 1)$, for instance. Moreover, let $xy \in
L (s_{xy})$ for $x, y \in \{ 0, 1 \}$. From the
initialization state $s$, the MDP moves to one of
the four internal states randomly with the uniform
distribution. Consider the internal state $s_{01}$. If the actions $a[a]$
or $a[b]$ are chosen, they denote that the event type $a$ is currently
observed. The MDP moves to the internal state $s_{01}$
and $s_{11}$ with probability $.375 = \frac{3}{8}$ and $.625 =
\frac{5}{8}$ respectively. Other transitions are defined similarly.

On the other hand, the initialization state $t$
corresponds to the computation on neighbors. The MDP moves to the internal states $t_{00}$,
$t_{01}$, $t_{10}$, $t_{11}$ randomly with
the uniform distribution. Similarly, $xy \in L (t_{xy})$ for $x, y \in
\{ 0, 1 \}$.
Consider, for example, the internal state
$t_{10}$. If the action $a[a]$ or $b[a]$ are chosen, they
denote that the current event type is replaced with the eveny type
$a$. The MDP moves to the internal state
$t_{00}$ and $t_{10}$ with probability $.375 =
\frac{3}{8}$ and $.625 = \frac{5}{8}$ respectively. At an internal
state, the MDP can move to the output states $0$,
$1$, or $2$ by the truncated $\frac{1}{2}$-geometric mechanism over
$\{ 0, 1, 2 \}$ (Figure~\ref{figure:randomized-frequency}, \`{a} la
Figure~\ref{figure:geometric-mechanism}). For any action sequence,
observe that paths from the state $s$ and those from $t$
formalize the computation of Algorithm~\ref{algorithm:frequency-estimator}
on an observation and its neighbor represented by the sequence. We
hence say the states $s$ and $t$ are neighbors in the
MDP (Figure~\ref{figure:frequency-estimator}).

\begin{figure}[tbh]
  \centering
  \begin{subfigure}{\columnwidth}
    \centering
      \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node
      distance=2cm,node/.style={circle,draw}]
      \node[node] (I) at  ( 0,  0) { $s$ };
      \node[node] (00) at (-1.5,  1.5) { $s_{00}$ };
      \node at (-1.9, 1.9) { $00$ };
      \node[node] (01) at (-1.5, -1.5) { $s_{01}$ };
      \node at (-1.9, -1.9) { $01$ };
      \node[node] (10) at ( 1.5,  1.5) { $s_{10}$ };
      \node at ( 1.9,  1.9) { $10$ };
      \node[node] (11) at ( 1.5, -1.5) { $s_{11}$ };
      \node at ( 1.9, -1.9) { $11$ };

@@

      \hspace{.05\columnwidth}

@@

      \end{tikzpicture}
      \caption{Randomized Frequency}
      \label{figure:randomized-frequency}
    \end{subfigure}
  \caption{Frequency Estimator (some states are drawn twice).}
  \label{figure:frequency-estimator}
\end{figure}

\noindent
\emph{Privacy Specification}.
For classical differential privacy, we are interested in the output
distribution. Algorithm~\ref{algorithm:frequency-estimator} has three
different outputs. By privacy analysis, the state $I$ satisfies the
following $\dpCTLstar$ formula:
$
\Psi = \dpriv{\ln 4}{0} (\F \mathit{at}_0) \wedge \dpriv{\ln 4}{0} (\F
\mathit{at}_1) \wedge \dpriv{\ln 4}{0} (\F \mathit{at}_2)
$.
For pan privacy, we moreover wish to measure privacy loss in internal
states. For observations of length $k$, the probabilities of reaching
each internal states need to be sufficiently close. The state $I$
therefore must satisfy
\[
\Psi
\wedge
\bigwedge_{k \in \bbfN}
\dpriv{\ln 4}{0} \X^k {00} \wedge
\dpriv{\ln 4}{0} \X^k {01} \wedge
\dpriv{\ln 4}{0} \X^k {10} \wedge
\dpriv{\ln 4}{0} \X^k {11}.
\]

### 2018-01-30 introduction.tex
remove notes
work through and 16 pages ;)

An important byproduct of this work is to analyze privacy over
infinite behaviors. Consider the formula $\dpriv{\epsilon}{\delta}
(\G \F \mathit{high})$. A state satisfies the formula if it
is probabilistically similar to always almost surely
answer $\mathit{high}$ as similar states. Compared to the unbounded
but finite behavior specified by $\F \mathit{high}$,
$\G \F \mathit{high}$ specifies an infinite behavior.
Not only can differential privacy specify randomized algorithms, it
also naturally generalizes to reactive systems.
Privacy analysis applies to systems with infinite behaviors as well.

@@

In order to verify $\dpCTLstar$ properties, we extend the standard
probabilistic model checking algorithms. For Markov chains, the states
satisfying a subformula $\dpriv{\epsilon}{\delta} \phi$ are computed
by a simple variant of the algorithm for Markov chains.
The time complexity of our model checking algorithm is the same as
those of $\CTLstar$ for Markov chains. The logic $\dpCTLstar$ obtains its
expressiveness for free. 
For Markov decision processes, we show that checking whether a state
satisfies $\dpriv{\epsilon}{\delta} \phi$ is undecidable.

@@

\paragraph{Contributions.} Our main contributions are three folds. (i)
We introduce the logic $\dpCTLstar$ for reasoning about properties in 
differential privacy. We use the logic to express classical
differential privacy properties, and moreover, complex properties such as those 
involving infinite behaviors can also be expressed. (ii) We model several differential privacy mechanisms, including the streaming process, in Markov chains or MDPs.
(iii) We show that the model checking problem 
problem for Markov chains is standard. For MDPs we show that it is undecidable.

>>>>

An important byproduct of this work is to analyze privacy over
infinite behaviors. Consider the formula $\dpriv{\epsilon}{\delta}
(\G \F \mathit{high})$. A state satisfies the formula if it
is probabilistically similar to always almost surely\lz{infinitely often}
answer $\mathit{high}$ as similar states. Compared to the unbounded
but finite behavior specified by $\F \mathit{high}$,
$\G \F \mathit{high}$ specifies an infinite behavior.
Not only can differential privacy specify randomized algorithms, it
also naturally generalizes to reactive systems.
Privacy analysis applies to systems with infinite behaviors as well.

@@

In order to verify $\dpCTLstar$ properties, we extend the standard
probabilistic model checking algorithms. For Markov chains, the states
satisfying a subformula $\dpriv{\epsilon}{\delta} \phi$ are computed
by a simple variant of the algorithm for Markov chains.
The time complexity of our model checking algorithm is the same as
those of $\PCTLstar$ for Markov chains. The logic $\dpCTLstar$ obtains its
expressiveness for free.
For Markov decision processes, we show that checking whether a state
satisfies $\dpriv{\epsilon}{\delta} \phi$ is undecidable.

@@

\paragraph{Contributions.} Our main contributions are three folds. (i)
We introduce the logic $\dpCTLstar$ for reasoning about properties in
differential privacy. We use the logic to express classical
differential privacy properties, and moreover, complex properties such as those
involving infinite behaviors can also be expressed. (ii) We model several differential privacy mechanisms, including the streaming process, in Markov chains or Markov decision processes.
(iii) We show that the model checking problem
problem for Markov chains is standard. For MDPs we show that it is undecidable.

### 2018-01-30 mdp.tex
work through and 16 pages ;)

\subsection{Markov Decision Processes}

@@

As we describe earlier, the user has the possibility of querying multiple mechanisms.
We use Markov decision processes to model such
interactive mechanisms. Specifically, external inputs are modeled by
actions. Behaviors associated with different inputs are modeled by
distributions associated with actions.

@@

\subsection{Semantics}
The logic $\dpCTLstar$ can be interpreted over MDPs. Semantics over Markov chains generalizes to MDPs as well. Note that two states with different enabled actions
are trivially distinguishable; no privacy can be preserved.
Let $M = (S, \Act, \wp, L)$ be an MDP. Define the satisfaction
relation $M, \neighbor{S}, s \models \Phi$ for $\PJ{J}$ and $\dpriv{\epsilon}{\delta} \phi$ (others are standard) as follows.
\begin{eqnarray*}
%  M, \neighbor{S}, s \models \top\\
%  M, \neighbor{S}, s \models p  & \textmd{ if } &  p \in L(s)\\
%  M, \neighbor{S}, s \models \neg \Phi  & \textmd{ if } &  M, \neighbor{S}, s \not\models \Phi\\
%  M, \neighbor{S}, s \models \Phi_0 \wedge \Phi_1  & \textmd{ if } &  M, \neighbor{S}, s \models \Phi_0 \textmd{ and }
%  M, \neighbor{S}, s \models \Phi_1\\
  M, \neighbor{S}, s \models \PJ{J} \phi
  & \textmd{ if } &
  \myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} \in J
  \textmd{ for every scheduler } \scheduler{S}\\
  M, \neighbor{S}, s \models \dpriv{\epsilon}{\delta} \phi
  & \textmd{ if } &
  \textmd{for all } t \textmd{ with } s \neighbor{S} t \textmd{ and
   scheduler } \scheduler{S},
   \myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} \leq
  e^{\epsilon} \cdot
   \\
  & &
   \myPr{t}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} + \delta
   \textmd{ and }
   \myPr{t}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} \leq e^{\epsilon} \cdot
   \myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} + \delta
\end{eqnarray*}

@@

\begin{theorem}
The $\dpCTLstar$ model checking problem for MDPs  is undecidable.
\end{theorem}

>>>>

%\subsection{Markov Decision Processes}

@@

As we describe earlier, the user has the possibility of querying multiple mechanisms.
We use MDPs to model such
interactive mechanisms. Specifically, external inputs are modeled by
actions. Behaviors associated with different inputs are modeled by
distributions associated with actions.

@@

\subsection{Semantics}
The logic $\dpCTLstar$ can be interpreted over MDPs. Semantics over Markov chains generalizes to MDPs as well. %Note that two states with different enabled actions
%are trivially distinguishable; no privacy can be preserved.
Let $M = (S, \Act, \wp, L)$ be an MDP. Define the satisfaction
relation $M, \neighbor{S}, s \models \Phi$ for $\PJ{J}$ and $\dpriv{\epsilon}{\delta} \phi$ (others are standard) as follows.
\begin{eqnarray*}
%  M, \neighbor{S}, s \models \top\\
%  M, \neighbor{S}, s \models p  & \textmd{ if } &  p \in L(s)\\
%  M, \neighbor{S}, s \models \neg \Phi  & \textmd{ if } &  M, \neighbor{S}, s \not\models \Phi\\
%  M, \neighbor{S}, s \models \Phi_0 \wedge \Phi_1  & \textmd{ if } &  M, \neighbor{S}, s \models \Phi_0 \textmd{ and }
%  M, \neighbor{S}, s \models \Phi_1\\
  M, \neighbor{S}, s \models \PJ{J} \phi
  & \textmd{ if } &
  \myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} \in J
  \textmd{ for every scheduler } \scheduler{S}\\
  M, \neighbor{S}, s \models \dpriv{\epsilon}{\delta} \phi
  & \textmd{ if } &
  \textmd{for all } t \textmd{ with } s \neighbor{S} t \textmd{ and
   scheduler } \scheduler{S},
   \myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} \leq
  e^{\epsilon} \cdot
   \\
  & &
   \myPr{t}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} + \delta
   \textmd{ and }
   \myPr{t}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} \leq e^{\epsilon} \cdot
   \myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} + \delta
\end{eqnarray*}

@@

\begin{theorem}\label{theorem:mdp-model-checking}
The $\dpCTLstar$ model checking problem for MDPs  is undecidable.
\end{theorem}

### 2018-01-30 syntax.tex
some changes
remove notes
work through and 16 pages ;)

Consider
\begin{eqnarray*}
  \Phi & ::= & \top \ |\ p \ |\ \neg \Phi \ |\ \Phi \wedge \Phi \ |\
               \PJ{J} \phi \ |\ \dpriv{\epsilon}{\delta} \phi\\
  \phi & ::= & \Phi \ |\ \neg\phi  \ |\  \phi\wedge\phi  \ |\  \X \phi \ |\ \phi \until \phi
\end{eqnarray*}
A \emph{state} formula $\Phi$ is either $\top$, an atomic proposition
$p$, the negation of a sate formula, the conjunction of two state
formulae, a \emph{probabilistic} operator $\PJ{J}$ with $J$
an interval in $[0, 1]$ followed by a path formula, or a
\emph{differentially private} operator $\dpriv{\epsilon}{\delta}$
 with two non-negative real numbers $\epsilon$ and $\delta$ followed
 by a path formula. A
\emph{path} formula $\phi$ is simply a LTL formula, with temporal operator \emph{next}  ($\X$)
and  \emph{until} operator
($\until$) operator enclosed by two path formulae.

As in the classical setting, we consider the sublogic $\dpCTL$ by allowing only path formulae of the form $\X\Phi$ and $\Phi\until\Phi$.

>>>>

The syntax of $\dpCTL$ state and path formulae are given by:
\begin{eqnarray*}
  \Phi & ::= & p \ |\ \neg \Phi \ |\ \Phi \wedge \Phi \ |\
               \PJ{J} \phi \ |\ \dpriv{\epsilon}{\delta} \phi\\
  \phi & ::= & \Phi \ |\ \neg\phi  \ |\  \phi\wedge\phi  \ |\  \X \phi \ |\ \phi \until \phi
\end{eqnarray*}
A \emph{state} formula $\Phi$ is either  an atomic proposition
$p$, the negation of a sate formula, the conjunction of two state
formulae, a \emph{probabilistic} operator $\PJ{J}$ with $J$
an interval in $[0, 1]$ followed by a path formula, or a
\emph{differentially private} operator $\dpriv{\epsilon}{\delta}$
 with two non-negative real numbers $\epsilon$ and $\delta$ followed
 by a path formula. A
\emph{path} formula $\phi$ is simply a LTL formula, with temporal operator \emph{next}  ($\X$)
and  \emph{until} operator
($\until$) operator enclosed by two path formulae.
We often use syntactic sugar $F$ and $G$ defined by: $F \phi:= true \until \phi$ and $G\phi :=\neg F (\neg\phi)$.

As in the classical setting, we consider the sublogic $\dpCTL$ by allowing only path formulae of the form $\X\Phi$ and $\Phi\until\Phi$.
Moreover, one obtains $\PCTL$ and $\PCTLstar$ from $\dpCTL$ and $\dpCTLstar$
by removing the differentially private operator  $\dpriv{\epsilon}{\delta}$.
